Dependent Object Types

### 2012-09-06 doc/candidacy/slides.tex
minor
red
punctuation

\begin{frame}{Limitations of SML modules}
\begin{itemize}
\item modules are not first-class; e.g. not possible to return a
  structure from an {\tt if} expression.
\item type sharing is restricted to equality between type names, not
  general type expressions; in effect, transparency / opaqueness of a
  signature cannot be fully fined-tuned
\end{itemize}
\end{frame}

\begin{frame}{ML: Translucent Sums}
\begin{itemize}
\item translucent sums are values representing modules
\item translucent sum: sequence of bindings
\item translucent sum type: sequence of declarations
\item unlike traditional records, later fields can depend on earlier ones
\item any type can be partially or fully determined by a type
\item dependent typing since translucent sums are terms that may contain type components
\end{itemize}
\end{frame}

\begin{frame}[fragile]{ML: Example}
\begin{minted}{sml}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
signature XType = sig
  type T
  val v = T
  val f = T -> T
end
\end{minted}
\end{frame}

@@

\begin{frame}{ML: Path-Dependent Types}
\begin{itemize}
\item Scala's abstract types  $\approx$ ML's abstract types of signatures; limitations:
\begin{itemize}
\item recursive references
\item bounded quantifications
\end{itemize}
\item translucent sum can be given a more precise type by referring to the name of its member path with a path selection; essential for
\begin{itemize}
\item breaking the dependencies between (sub)fields
\item propagation of typing information
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{Tribe Calculus: Virtual Classes}
\begin{itemize}
\item {\it familes} of classes inherited together
\item classes are lexically nested inside other classes: when a class
  is inherited, its nested inner classes are inherited along with
  their methods and fields
\item two form of inheritance: subclassing and further binding
\item {\it family polymorphism}: code written for one family
  also works for extensions of that family
\item in Tribe:\begin{itemize}
\item path types can depend simultaneously on both classes and objects
\item paths can use an {\tt out} field to move from an object to the
  object which surrounds it; enables ubiquitous access to an object's
  family without the need to drag around family arguments.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Tribe Types}
\begin{itemize}
\item In Tribe, types are a form of generalized paths: a final
  variable, {\tt this} or a class name then followed by a possibly
  empty sequence of field, {\tt out} or class selections.
\item {\tt kitt.Passenger.name}: the name of one of Kitt's passengers
\item {\tt kitt.driver <: Car.driver <: Car.Passenger <: Car.Traveller <:
  Vehicle.Traveller}.
\item {\tt kitt.driver $\not<:$ karr.driver} and {\tt kitt.Passenger $\not<:$ karr.Passenger}.
\end{itemize}
\begin{minted}[frame=lines,fontsize=\footnotesize]{java}
class Vehicle { class Traveller { ... } }
class Car extends Vehicle {
  class Passenger extends Traveller {
    class String { ... }
    final String name;
  }
  final Passenger driver;
}
final Car kitt;
final Car karr;
\end{minted}
\end{frame}

>>>>

\begin{frame}{Limitations of SML modules}
\begin{itemize}
\item modules are not first-class; e.g. not possible to return a
  structure from an {\tt if} expression
\item type sharing is restricted to equality between type names, not
  general type expressions; in effect, transparency / opaqueness of a
  signature cannot be fully fined-tuned
\end{itemize}
\end{frame}

\begin{frame}{ML-Like: Translucent Sums}
\begin{itemize}
\item translucent sums are values representing modules
\item translucent sum: sequence of bindings
\item translucent sum type: sequence of declarations
\item unlike traditional records, later fields can depend on earlier ones
\item any type can be partially or fully determined by a type expression
\item dependent typing since translucent sums are terms that may contain type components
\end{itemize}
\end{frame}

\begin{frame}[fragile]{ML-Like: Example}
\begin{minted}{sml}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
signature XType = sig
  type T
  val v = T
  val f = T -> T
end
\end{minted}
\end{frame}

@@

\begin{frame}{ML-Like: Path-Dependent Types}
\begin{itemize}
\item Scala's abstract types  $\approx$ ML's abstract types of signatures; limitations:
\begin{itemize}
\item recursive references
\item bounded quantifications
\end{itemize}
\item translucent sum can be given a more precise type by referring to the name of its member path with a path selection; essential for
\begin{itemize}
\item breaking the dependencies between (sub)fields
\item propagation of typing information
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{Tribe Calculus: Virtual Classes}
\begin{itemize}
\item {\it familes} of classes inherited together
\item classes are lexically nested inside other classes: when a class
  is inherited, its nested inner classes are inherited along with
  their methods and fields
\item two form of inheritance: subclassing and further binding
\item {\it family polymorphism}: code written for one family
  also works for extensions of that family
\item in Tribe:\begin{itemize}
\item path types can depend simultaneously on both classes and objects
\item paths can use an {\tt out} field to move from an object to the
  object which surrounds it; enables ubiquitous access to an object's
  family without the need to drag around family arguments
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Tribe Types}
\begin{itemize}
\item in Tribe, types are a form of generalized paths: a final
  variable, {\tt this} or a class name then followed by a possibly
  empty sequence of field, {\tt out} or class selections
\item {\tt kitt.Passenger.name}: the name of one of Kitt's passengers
\item {\tt kitt.driver <: Car.driver <: Car.Passenger <: Car.Traveller <:
  Vehicle.Traveller}
\item {\tt kitt.driver $\not<:$ karr.driver} and {\tt kitt.Passenger $\not<:$ karr.Passenger}
\end{itemize}
\begin{minted}[frame=lines,fontsize=\footnotesize]{java}
class Vehicle { class Traveller { ... } }
class Car extends Vehicle {
  class Passenger extends Traveller {
    class String { ... }
    final String name;
  }
  final Passenger driver;
}
final Car kitt;
final Car karr;
\end{minted}
\end{frame}
